---
title: "Elephant Seal Age JAE Intraspecific Variation"
format: 
  html: 
    echo: false
    fig-width: 5
    fig-height: 5
    code-fold: true
    code-tools: true
    warning: false
editor: visual
bibliography: references.bib
---

```{r}
#| label: setup
#| include: false  
library(ggthemes)
library(cowplot)
library(lme4)
library(lmerTest) # Loading this library will automatically append p-values to lmer models estimated using something called Satterthwaite's method
library(merTools)
library(tidyverse)
library(ggeffects)
library(ggridges)
library(patchwork)
source(here::here("R/format.R"))
theme_set(theme_few(base_size = 13))

age_senesce <- 11

sealdat <- read_csv(here::here("data/raw/128L pull 2023_12_05.csv"), 
                    show_col_types = FALSE) %>% 
  mutate(observed = if_else(observed == "B", "Breeder", "Non-breeder"), 
         observed_int = if_else(observed == "Breeder", 1, 0), 
         pup_survived = pupseeneveragain > 0) %>% 
  filter(age > 3, 
         year > 1987) %>% 
  mutate(animalID = factor(animalID),
         year_fct = factor(year),
         age10 = age / 10,
         age_cat = factor(age >= age_senesce, 
                          labels = c("Young", "Old"))) %>% 
  group_by(animalID) %>% 
  mutate(longevity = ifelse(max(year) < 2020,
                            max(age),
                            NA),
         longevity10 = longevity / 10) %>% 
  ungroup()

resight <- read.csv(here::here("data/raw/fullresights.csv")) %>% 
  select(animalID, resightID, season, date, timeofyear, calyear, yday) %>% 
  mutate(animalID = as.factor(animalID))

```

# Introduction

Our hypotheses:

1.  Elephant seal reproductive success declines with age.

2.  Reproductive declines persist to the next generation (maternal effect senescence). That is, offspring survival and recruitment are decreased.

3.  Offspring sex ratios shift towards males with old age.

4.  Phenology is shifted later for older animals and breeders.

# Results

First, address detection probability. DO THIS IN MARK.

n = `r n_distinct(sealdat$animalID)` distinct animals.

### Raw data

Resighting plot (image) and age structure bar plot

```{r}
#| label: fig-raw-data

sighting_img <- sealdat %>% 
  group_by(animalID) %>% 
  summarize(yearborn = yearborn[1],
            first_obs = min(year),
            last_obs = max(year),
            .groups = "keep") %>% 
  reframe(yearborn = yearborn,
          year = yearborn:last_obs,
          age = year - yearborn,
          prerecruit = year < first_obs) %>% 
  filter(year > 1987) %>% 
  left_join(sealdat %>% 
              group_by(animalID, year) %>% 
              summarize(observed_int = max(observed_int),
                        .groups = "drop"),
            by = c("animalID", "year")) %>% 
  mutate(
    observed = case_when(
      age == 0 ~ "Born",
      prerecruit ~ "Pre-recruit",
      observed_int == 0 ~ "Non-breeder",
      observed_int == 1 ~ "Breeder",
      is.na(observed_int) ~ "Unobserved"
    ),
    observed = factor(observed, levels = c("Born",
                                           "Pre-recruit",
                                           "Non-breeder",
                                           "Breeder",
                                           "Unobserved"))
  ) %>% 
  group_by(animalID) %>% 
  mutate(longevity = max(age),
         yearborn = na.omit(yearborn)[1]) %>% 
  ungroup() %>% 
  arrange(yearborn, longevity, animalID) %>% 
  mutate(animalID = factor(animalID, levels = unique(animalID))) %>% 
  ggplot(aes(year, animalID)) +
  geom_raster(aes(fill = observed)) +
  scale_fill_viridis_d(direction = -1) +
  guides(fill = guide_legend(position = "inside", 
                             direction = "horizontal",
                             ncol = 1,
                             byrow = TRUE)) +
  labs(y ="Individual") +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.x = element_blank(),
        legend.title = element_blank(),
        legend.position.inside = c(0.01, 0.99),
        legend.justification = c(0, 1),
        legend.background = element_blank(),
        legend.key.size = unit(4, "mm"),
        legend.text = element_text(size = 8))

age_dist <- sealdat %>% 
  filter(age < 22) %>% 
  group_by(age) %>% 
  count(age) %>% 
  ggplot(aes(x = age, y = n)) + 
  geom_col(fill = "grey50") + 
  labs(x = "Age (Years)", y = "Number of individuals") + 
  theme()

(sighting_img | age_dist) + plot_layout(widths = c(2, 1))

```

### H1: Age distribution and breeding status

~~We predicted that breeding success would decline with age. We found that breeding success increases to prime age, then decreases at older ages (@fig-age_dist).~~

```{r}
#| label: breeding-mod

breed_age_mod <- glmer(
  observed_int ~ poly(age10, 2, raw = TRUE) + (1 | animalID) + (1 | year_fct),
  sealdat,
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

summary(breed_age_mod)

```

```{r}
#| label: fig-breeding-mod 
#| fig-width: 7.5

# Summarize the raw data
breed_nonbreed <- sealdat %>% 
  group_by(age) %>% 
  summarize(n_observed = n(), 
            perc_breed = sum(observed_int) / n_observed, 
            se = (perc_breed * (1 - perc_breed) / n_observed)^0.5) %>% 
  filter(age <= 20)

# Create POPULATION-LEVEL (i.e. without rand effects) predictions
breed_pred_pop <- ggpredict(
  breed_age_mod, 
  terms = "age10 [all]",
  type = "fixed",
  interval = "confidence"
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10) %>% 
  filter(age <= 20)

# Create WEIGHTED-AVERAGE predictions using random effects
seal_years <- count(sealdat, year_fct)
breed_pred_wgt <- ggpredict(
  breed_age_mod,
  terms = c("age10 [all]", "year_fct [all]"),
  type = "random",
  interval = "confidence"
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10,
         year_fct = group) %>% 
  filter(age <= 20) %>% 
  left_join(seal_years, by = "year_fct") %>% 
  group_by(age) %>% 
  summarize(predicted = weighted.mean(predicted, n))
  
ggplot(breed_pred_pop, aes(age, predicted)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), 
              alpha = 0.2) +
  geom_line(linewidth = 1.2, color = "grey50") +
  geom_line(data = breed_pred_wgt, 
            color = "cornflowerblue", 
            linewidth = 1.2) +
  geom_pointrange(aes(y = perc_breed, 
                      ymin = perc_breed - se, 
                      ymax = perc_breed + se),
                  breed_nonbreed) +
  scale_y_continuous("Breeding", labels = scales::percent) +
  labs(x = "Age (years)") +
  theme(legend.position = "none")

```

Sample size n = `r nrow(sealdat)`.

### H2: Maternal effect senescence

We predicted that reproductive declines would carry over to the next generation - that is, pups born to older mothers would have decreased survival and recruitment success.

We found that neither pup survival nor recruitment is significantly affected by maternal age (@fig-repro-age).

We reject our hypothesis: reproductive declines do not persist to the next generation, and there is no evidence of maternal effect senescence.

```{r}
#| label: repro-age

##NOTE: Need to take out the instances where recruitment > survival
#what does ggpredict do with random effects? 
#thin gray lines for the different year effects? show the interannual variation (5 that include extreme ones, just something to illustrate the variation of year)

# Isolate pup survival/recruitment 
surv_recruit_data <- sealdat %>% 
  drop_na(pupID) %>% 
  # Only pups born 2016 or earlier (otherwise survival biiased)
  filter(pupyearborn <= 2016) %>% 
  mutate(pup_survived_int = ifelse(pup_survived, 1L, 0L)) %>% 
  # Only keep maternal age classes with 5+ births, which drops moms 19+
  group_by(age) %>% 
  mutate(n_pups = n()) %>% 
  ungroup() %>%
  filter(n_pups >= 5)

# Fit pup *survival* models pre- and post-senescence
# Commented-out formulae vary age10 coef by pupsex, end up with similar
# results. Can add to supplemental?
pup_surv_model1 <- glmer(
  # pup_survived_int ~ age10 : pupsex + (1 | animalID) + (1 | year),
  pup_survived_int ~ age10 + (1 | animalID) + (1 | year),
  data = filter(surv_recruit_data, age < age_senesce),
  family = "binomial"
)
pup_surv_model2 <- glmer(
  # pup_survived_int ~ age10 : pupsex + (1 | animalID) + (1 | year),
  pup_survived_int ~ age10 + (1 | animalID) + (1 | year),
  data = filter(surv_recruit_data, age >= age_senesce),
  family = "binomial"
)

# Fit pup *recruitment* models pre- and post-senescence
pup_recr_model1 <- glmer(
  puprecruited ~ age10 + (1 | animalID) + (1 | year),
  # For recruitment, ONLY female pups
  data = filter(surv_recruit_data, 
                age < age_senesce,
                pupsex == "F"),
  family = "binomial"
)
pup_recr_model2 <- glmer(
  puprecruited ~ age10 + (1 | animalID) + (1 | year),
  data = filter(surv_recruit_data, 
                age >= age_senesce,
                pupsex == "F"),
  family = "binomial",
  control = glmerControl(optimizer="bobyqa")
)

# Summary table
surv_recr_summ <- tibble(
  response = rep(c("Survival", "Recruitment"), each = 2),
  n = map_dbl(
    list(pup_surv_model1@frame, 
         pup_surv_model2@frame, 
         pup_recr_model1@frame, 
         pup_recr_model2@frame),
    nrow
  ),
  model = rep(c("Pre-senescence", "Post-senescence"), 2),
  age_coef = c(fixef(pup_surv_model1)["age10"],
               fixef(pup_surv_model2)["age10"],
               fixef(pup_recr_model1)["age10"],
               fixef(pup_recr_model2)["age10"]),
  age_coef_se = c(summary(pup_surv_model1)$coef["age10", "Std. Error"],
                  summary(pup_surv_model2)$coef["age10", "Std. Error"],
                  summary(pup_recr_model1)$coef["age10", "Std. Error"],
                  summary(pup_recr_model2)$coef["age10", "Std. Error"]),
  `p(coef<0)` = c(pnorm(summary(pup_surv_model1)$coef["age10", "z value"]),
                  pnorm(summary(pup_surv_model2)$coef["age10", "z value"]),
                  pnorm(summary(pup_recr_model1)$coef["age10", "z value"]),
                  pnorm(summary(pup_recr_model2)$coef["age10", "z value"]))
)

surv_recr_summ %>% 
  mutate(across(where(is.numeric), \(x) signif(x, 3))) %>% 
  knitr::kable()

```

```{r}
#| label: fig-repro-age
#| fig-cap: "Offspring survival decreases as mothers senesce. **A** Survival of female and male offspring (filled points) is not negatively impacted by maternal age (blue line) prior to age 12 (vertical dashed line). After age 12, offspring survival declines with maternal age (red line), i.e., maternal effect senescence. Hollow points represent female offspring recruitment, which did not significantly decline with maternal age after age 12. **B** The coefficient of age w.r.t. offspring survival (solid points) was not significantly negative before age 12 (blue), but was significantly negative after (red). The coefficient of age w.r.t. female offspring recruitment (hollow points) was not significant in either maternal age class. Lines represent standard error of the coefficient, which is ommitted for the post-age 12 offspring recruitment coefficent (hollow red point) because the standard error was too large (14.8) to visually represent without hiding other patterns. Numbers at top indicate sample size in each model."
#| fig-width: 5.5

#B Pup survival and recruitment doesn't change with increased maternal age.

# Pup survival 
surv_pred <- rbind(
  # Pre-senescence
  ggpredict(
    pup_surv_model1,
    terms = c("age10 [all]"),
    type = "fixed",
    interval = "confidence"
  ) %>% 
    as_tibble() %>% 
    mutate(age = x * 10,
           age_class = "young"),
  # Post-senescence
  ggpredict(
    pup_surv_model2,
    terms = c("age10 [all]"),
    type = "fixed",
    interval = "confidence"
  ) %>% 
    as_tibble() %>% 
    mutate(age = x * 10,
           age_class = "old")
)

repro_raw <- surv_recruit_data %>% 
  # Uncomment to group by female and male offspring
  # mutate(
  #   pupsex = factor(pupsex, 
  #                   labels = c("Female offspring", "Male offspring"))
  # ) %>% 
  # group_by(age, pupsex) %>% 
  group_by(age) %>% 
  summarize(n_born = n(), 
            n_surv = sum(pup_survived_int),
            perc_surv = n_surv / n_born, 
            se_surv = (perc_surv * (1 - perc_surv) / n_born)^0.5,
            f_born = sum(pupsex == "F"),
            f_recr = sum(puprecruited & pupsex == "F"),
            perc_recr = f_recr / f_born,
            se_recr = (perc_recr * (1 - perc_recr) / f_born)^0.5,
            .groups = "drop") 

surv_recr_fig <- ggplot(surv_pred, aes(x = age)) +
  # Survival by age (fixed effect only)
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, 
                  group = age_class),
              fill = "grey", alpha = 0.5) +
  geom_line(aes(y = predicted, color = age_class), 
            linewidth = 1.5) +
  # Raw survival and recruitment data
  geom_pointrange(aes(y = perc_surv,
                      ymin = perc_surv - se_surv,
                      ymax = perc_surv + se_surv),
                  repro_raw) +
  geom_pointrange(aes(y = perc_recr,
                      ymin = perc_recr - se_recr,
                      ymax = perc_recr + se_recr),
                  repro_raw,
                  shape = 21) +
  # Add visual indicator for senescence
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_x_continuous("Age", limits = c(3.5, 18.5)) +
  scale_y_continuous("Offspring survival/recruitment", 
                     limits = c(0, 0.65), 
                     labels = scales::percent) +
  scale_color_manual(values = c("firebrick", "cornflowerblue")) +
  # Uncomment to facet by offspring sex
  # facet_wrap(~pupsex) +
  theme_few() +
  theme(legend.position = "none",
        axis.title = element_text(size = 13))

# Survival/recruitment figure. In the caption, mention the post-senescence
# recruitment std err was omitted because it was 14.8, so it swamps all the
# other std errs
surv_recr_coef_fig <- surv_recr_summ %>% 
  mutate(response = fct_rev(str_sub(response, 1, 4)),
         model = fct_rev(str_extract(model, ".*-")),
         # Remove post-senescence recruitment std err
         age_coef_se = c(age_coef_se[1:3], 0)) %>% 
  ggplot(aes(x = response, ymax = 1.1)) +
  geom_pointrange(aes(ymin = age_coef - age_coef_se, 
                      y = age_coef,
                      ymax = age_coef + age_coef_se,
                      color = model,
                      shape = response),
                  position = position_dodge(width = 0.9)) +
  geom_text(aes(label = n, group = model), 
            y = 0.9, 
            size = 3,
            angle = 30,
            hjust = 0,
            position = position_dodge(width = 1)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("cornflowerblue", "firebrick")) +
  scale_shape_manual(values = c(16, 1)) +
  labs(x = "Response", y = "Age coefficient") +
  expand_limits(y = 1.1) +
  theme(legend.position = "none")

(surv_recr_fig | surv_recr_coef_fig) + 
  plot_layout(widths = c(3, 1)) +
  plot_annotation(tag_levels = "A")

```

Sample size n = `r nrow(surv_recruit_data)`.

### H3: Changes in pup sex ratio with increased maternal age

We found that as maternal age increases, pup sex skews slightly towards males (@fig-sex_ratio).

```{r}
#| label: sex-ratio

pup_sex_data <- sealdat %>% 
  filter(pupsex %in% c("M", "F"), 
         age < 19) %>% 
  mutate(is_male = ifelse(pupsex == "M", 1, 0))

sex_ratio_mod1 <- glmer(
  is_male ~ age10 + (1 | animalID) + (1 | year),
  filter(pup_sex_data, age < age_senesce),
  family = "binomial"
)
sex_ratio_mod2 <- glmer(
  is_male ~ age10 + (1 | animalID) + (1 | year),
  filter(pup_sex_data, age >= age_senesce),
  family = "binomial"
)

sex_ratio_summ <- tibble(
  model = c("Pre-senescence", "Post-senescence"),
  age_coef = c(fixef(sex_ratio_mod1)["age10"],
               fixef(sex_ratio_mod2)["age10"]),
  age_coef_se = c(summary(sex_ratio_mod1)$coef["age10", "Std. Error"],
                  summary(sex_ratio_mod2)$coef["age10", "Std. Error"]),
  `p(coef!=0)` = c(summary(sex_ratio_mod1)$coef["age10", "Pr(>|z|)"],
                   summary(sex_ratio_mod2)$coef["age10", "Pr(>|z|)"])
) 

sex_ratio_summ %>% 
  mutate(across(where(is.numeric), \(x) signif(x, 3))) %>%
  set_names(c("Age class", 
              "Age", 
              "se(Age)",
              "p(Age) != 0")) %>% 
  knitr::kable()

```

```{r}
#| label: fig-sex-ratio
#| fig-cap: "The effect of age on offspring sex ratio was not significantly different for pre- and post-senescence adults. **A** Raw data (points with error bars, indicating mean and standard error of the proportion) and model predictions (lines with 95% confidence intervals). **B** Estimate and standard error of age coefficients in pre- and post-senescence models."
#| fig-width: 5.5

# Model predictions
sex_ratio_pred <- rbind(
  # Pre-senescence
  ggpredict(
    sex_ratio_mod1,
    terms = c("age10 [all]"),
    type = "fixed",
    interval = "confidence"
  ) %>% 
    as_tibble() %>% 
    mutate(age = x * 10,
           # Add an age_class variable to facilitate plotting
           age_class = "young"),
  # Post-senescence
  ggpredict(
    sex_ratio_mod2,
    terms = c("age10 [all]"),
    type = "fixed",
    interval = "confidence"
  ) %>% 
    as_tibble() %>% 
    mutate(age = x * 10,
           age_class = "old")
)

# Sex ratio raw data
sex_ratio <- pup_sex_data %>% 
  group_by(age) %>% 
  summarize(n_observed = n(), 
            perc_male = sum(is_male) / n_observed, 
            se = (perc_male * (1 - perc_male) / n_observed)^0.5)

# Figure with raw data and model predictions
sex_ratio_fig <- ggplot(sex_ratio_pred, aes(x = age)) +
  # Prediction across years (fixed effect only)
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, 
                  group = age_class),
              fill = "grey", alpha = 0.5) +
  geom_line(aes(y = predicted, color = age_class), 
            linewidth = 1.5) +
  # Raw data
  geom_pointrange(aes(y = perc_male, 
                      ymin = perc_male - se, 
                      ymax = perc_male + se),
                  sex_ratio) + 
  # Add visual indicator for senescence
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_y_continuous("Male offspring", 
                     labels = scales::percent) +
  scale_color_manual(values = c("firebrick", "cornflowerblue")) +
  labs(x = "Age") +
  theme_few() +
  theme(legend.position = "none",
        axis.title = element_text(size = 13))

# Coefficient panel
sex_ratio_coef_fig <- sex_ratio_summ %>% 
  mutate(model = str_extract(model, ".*-"),
         model = fct_rev(model)) %>% 
  ggplot(aes(color = model)) + 
  geom_pointrange(aes(ymin = age_coef - age_coef_se,
                      y = age_coef,
                      ymax = age_coef + age_coef_se,
                      x = model)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("cornflowerblue", "firebrick")) +
  labs(x = "Senescence",
       y = "Age coefficient") +
  theme_few() +
  theme(legend.position = "none")

(sex_ratio_fig | sex_ratio_coef_fig) + 
  plot_layout(widths = c(3, 1)) +
  plot_annotation(tag_levels = "A")
```

Sample size n = `r nrow(pup_sex_data)`.

There appears to be a step change biasing offspring sex ratio towards males post-senescence. We simulated the sex ratio of 1,000,000 offspring born to mothers pre- and post-senesecence using the coefficient estimates and standard errors of the fitted sex ratio models. The mean offspring sex ratio increases from 51.3% male before maternal age 12 to 61.2% male after maternal age 12. However, the uncertainty in post-senescence sex ratio is considerable, and only 59.3% of the contrasts of simulated post- and pre-senescence sex ratios are biased towards male offspring. I.e., a simulated post-senescent adult female was more likely to have a male offspring than a simulated pre-senescent adult female in 59.3% of simulations.

```{r}
#| label: fig-sim-sex-ratio
#| fig-cap: "Simulated sex ratios of offspring born to pre- and post-senescence adults. **A** Comparison of simulated sex ratios by pre- (blue) and post- (red) senescence adults. The mean (vertical lines) sex ratio of post-senescence adults is greater than pre-senescence adults. However the distributions overlap extensively. **B** The contrast of post-senescence sex ratio relative to pre-senescence is greater than 0 in only 59.3% of simulations."
#| fig-width: 5.5

# pre_sex_ratio <- tibble(age10 = seq(4, age_senesce - 1)) %>% 
inv_logit <- \(x) exp(x) / (1 + exp(x))
n_sim <- 1e6
sim_sex_ratio <- rbind(
  tibble(intercept_mean = summary(sex_ratio_mod1)$coef[1, 1],
         intercept_sd = summary(sex_ratio_mod1)$coef[1, 2],
         slope_mean = summary(sex_ratio_mod1)$coef[2, 1],
         slope_sd = summary(sex_ratio_mod1)$coef[2, 2],
         age10 = sample(seq(4, age_senesce - 1) / 10, n_sim, TRUE),
         intercept = rnorm(n_sim, intercept_mean, intercept_sd),
         slope = rnorm(n_sim, slope_mean, slope_sd),
         logit_p = intercept + slope * age10,
         p = inv_logit(logit_p),
         age_class = "Pre-senescence"),
  tibble(intercept_mean = summary(sex_ratio_mod2)$coef[1, 1],
         intercept_sd = summary(sex_ratio_mod2)$coef[1, 2],
         slope_mean = summary(sex_ratio_mod2)$coef[2, 1],
         slope_sd = summary(sex_ratio_mod2)$coef[2, 2],
         age10 = sample(seq(age_senesce, 18) / 10, n_sim, TRUE),
         intercept = rnorm(n_sim, intercept_mean, intercept_sd),
         slope = rnorm(n_sim, slope_mean, slope_sd),
         logit_p = intercept + slope * age10,
         p = inv_logit(logit_p),
         age_class = "Post-senescence")
) %>% 
  mutate(age_class = fct_rev(age_class))

pre_logit_p <- with(sim_sex_ratio, logit_p[age_class == "Pre-senescence"])
post_logit_p <- with(sim_sex_ratio, logit_p[age_class == "Post-senescence"])
sex_ratio_contrast <- tibble(
  contrast_logit_p = post_logit_p - pre_logit_p
)

sim_sex_ratio_fig <- ggplot(sim_sex_ratio, aes(p)) +
  geom_density(aes(fill = age_class), alpha = 0.5) +
  geom_density(aes(color = age_class)) +
  geom_vline(aes(xintercept = p, color = age_class),
             sim_sex_ratio %>% 
               group_by(age_class) %>% 
               summarize(p = inv_logit(mean(logit_p))),
             linewidth = 1.5) +
  scale_x_continuous("Male offspring", 
                     labels = scales::percent) +
  scale_color_manual(values = c("cornflowerblue", "firebrick")) +
  scale_fill_manual(values = c("cornflowerblue", "firebrick")) +
  theme(legend.position = "none")

sim_sr_contrast_fig <- ggplot(sex_ratio_contrast, aes(contrast_logit_p)) +
  geom_density(fill = "grey90") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "Contrast of logit(p)")

(sim_sex_ratio_fig | sim_sr_contrast_fig) +
  plot_annotation(tag_levels = "A")
  
```

### Supplemental Material

```{r}
#| label: seg-breeding-mod

# Fit GLMMs to the data with binary variable indicating before and after onset # of senescence. We a priori chose age 11 for senescence (age_senesce).
breed_age_seg <- glmer(
  observed_int ~ age10 * age_cat + (1 | animalID) + (1 | year_fct),
  sealdat,
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

summary(breed_age_seg)
```

```{r}
#| label: fig-seg-breeding-mod

breed_pred_seg <- ggpredict(breed_age_seg, 
                            terms = c("age10 [all]", "age_cat")) %>% 
  as_tibble() %>% 
  mutate(age = x * 10,
         age_cat = factor(group, levels = c("Young", "Old"))) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce))

ggplot(breed_pred_seg, aes(age, predicted)) +
  geom_ribbon(aes(fill = age_cat, ymin = conf.low, ymax = conf.high), 
              alpha = 0.2) +
  geom_line(aes(color = age_cat)) +
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_color_manual(values = c("cornflowerblue", "firebrick")) +
  scale_fill_manual(values = c("cornflowerblue", "firebrick")) +
  scale_y_continuous("Breeding", labels = scales::percent) +
  labs(x = "Age (years)") +
  theme(legend.position = "none")

```

Supplemental Table 1: Statistics for breeding success model.

```{r, code = FALSE}
#| label: tbl-breed-cap

tbl_breed_cap <- str_glue("The results from breeding probability analysis. The random effect variance of animal id is {format_signif(VarCorr(breeding_mod)$animalID, 3)} and of year was {format_signif(VarCorr(breeding_mod)$year, 3)}.")

```

```{r, code = FALSE}
#| label: tbl-breed-success
tbl-cap: !expr "tbl_breed_cap"

breed_mod_summary$coefficients %>% 
  as_tibble() %>% 
  mutate(Parameter = c("Intercept", "Age (linear)", "Age (quadratic)"),
         across(c(Estimate, `Std. Error`, `z value`), 
                \(x) format_signif(x, digits = 3)),
         `p value` = format_pval(`Pr(>|z|)`, digits = 2)) %>%
  select(-`Pr(>|z|)`) %>% 
  relocate(Parameter) %>% 
  knitr::kable()
```

```{r, code = FALSE}
#| label: tbl-survrecruit-cap

surv_recruit_cap <- str_glue("The results from offspring survival and recruitment analyses. In the offspring survival model, the random effect variance of animal id is {format_signif(VarCorr(pup_surv_model)$animalID, 3)} and of year was {format_signif(VarCorr(pup_surv_model)$year, 3)}. In the offspring recruitment model, the random effect variance of animal id is {format_signif(VarCorr(pup_recruit_model)$animalID, 3)} and of year was {format_signif(VarCorr(pup_recruit_model)$year, 3)}")

```

Supplemental table for offspring survival

```{r, code = FALSE}
#| label: tbl-surv_recruit
tbl-cap: !expr "surv_recruit_cap"

surv_mod_summary <- summary(pup_surv_model)
surv_mod_kable <- surv_mod_summary$coefficients %>% 
  as_tibble() %>% 
  mutate(Parameter = c("Intercept", "Age (linear)", "Age (quadratic)"),
         across(c(Estimate, `Std. Error`, `z value`), 
                \(x) format_signif(x, digits = 3)),
         `p value` = format_pval(`Pr(>|z|)`, digits = 2),
         Response = "Survival") %>%
  select(-`Pr(>|z|)`) %>% 
  relocate(Response, Parameter)

recruit_mod_summary <- summary(pup_recruit_model)
recruit_mod_kable <- recruit_mod_summary$coefficients %>% 
  as_tibble() %>% 
  mutate(Parameter = c("Intercept", "Age (linear)", "Age (quadratic)"),
         across(c(Estimate, `Std. Error`, `z value`), 
                \(x) format_signif(x, digits = 3)),
         `p value` = format_pval(`Pr(>|z|)`, digits = 2),
         Response = "Recruitment") %>%
  select(-`Pr(>|z|)`) %>% 
  relocate(Response, Parameter)

rbind(surv_mod_kable, recruit_mod_kable) %>% 
  knitr::kable() %>% 
  kableExtra::kable_styling() %>% 
  kableExtra::collapse_rows(columns = 1)


```

Supplemental table for pup sex ratio model.

```{r, code = FALSE}
#| label: tbl-pupsex-cap

pupsex_cap <- str_glue("The results from offspring sex ratio analyses. In the model, the random effect variance of animal id is {format_signif(VarCorr(pup_sex_mod)$animalID, 3)} and of year was {format_signif(VarCorr(pup_sex_mod)$year, 3)}.") 
```

```{r, code = FALSE}
#| label: tbl-pupsex
tbl-cap: !expr "pupsex_cap"

pupsex_mod_summary <- summary(pup_sex_mod)
pupsex_quad_summary <- summary(pup_sex_quad)
pupsex_mod_summary$coefficients %>% 
  as_tibble() %>% 
  mutate(Parameter = c("Intercept", "Age (linear)"),
         across(c(Estimate, `Std. Error`, `z value`), 
                \(x) format_signif(x, digits = 3)),
         `p value` = format_pval(`Pr(>|z|)`/2, digits = 3)) %>%
  select(-`Pr(>|z|)`) %>% 
  relocate(Parameter) %>% 
  knitr::kable()


# We fit a quadratic equation, but only the linear term was significant, so the linear model is the one that is plotted. take the p = 0.517 and divide by 2 because a priori. report in supplemental table that its reporting on 1 tailed (no absolute value of z, just z)
```

Supplemental table for male biased sex ratios at older ages.

```{r, code = FALSE}
#| label: tbl-pup-sex
#| tbl-cap: "Offspring sex ratio skewed towards males with increasing maternal age. Mothers age 12, 13, and 17 gave birth to significantly more male than female pups (bold)."

pup_sex_tbl <- pup_sex_summary %>% 
  transmute(
    Age = age,
    `Male pups` = M,
    `Total pups` = num,
    `% male` = scales::label_percent(accuracy = 0.1)(percm),
    `p value` = format_pval(pval, 3)
  )

is_signif <- which(pup_sex_tbl$`p value` < "0.05")

knitr::kable(pup_sex_tbl) %>% 
  kableExtra::kable_styling() %>% 
  kableExtra::row_spec(is_signif, bold = TRUE)
```

Supplemental figure for phenology models

```{r, code = FALSE}
#| fig-width: 8

#NOTE: Both BA and BD have some points that are really late (above 300 doy) that need fixing. Maybe make boxplots also..?
seal_phenology %>% 
  rename(`Breeding\narrival` = BA,
         `Breeding\ndeparture` = BD,
         `Molt\narrival` = MA,
         `Molt\ndeparture` = MD) %>% 
  pivot_longer(`Breeding\narrival`:`Molt\ndeparture`, 
               names_to = "event", 
               values_to = "DOY") %>% 
ggplot(aes(y = DOY)) +
  geom_boxplot(aes(factor(age))) +
  geom_smooth(aes(as.integer(factor(age))), method = "lm", formula = y ~ x, se = TRUE) +
  facet_wrap(~ event, nrow = 2, scales = "free_y") +
  labs(x = "Age (years)", y = "Day of year") +
  theme_classic()
```

Supplemental table for the phenology models.

```{r, code = FALSE}
ba_data <- phenology_models$model_data[[1]]
ba_mod <- lmer(doy ~ age + (1 | animalID) + (1 | year), data = ba_data)
ba_summary <- summary(ba_mod)
ba_ktable <- ba_summary$coefficients %>% 
  as_tibble() %>% 
  mutate(Parameter = c("Intercept", "Age (linear)"),
         across(c(Estimate, `Std. Error`, `df`, `t value`), 
                \(x) format_signif(x, digits = 3)),
         `p value` = format_pval(`Pr(>|t|)`, digits = 2),
         Response = "Breeding arrival") %>%
  select(-`Pr(>|t|)`) %>% 
  relocate(Response, Parameter)

bd_data <- phenology_models$model_data[[2]]
bd_mod <- lmer(doy ~ age + (1 | animalID) + (1 | year), data = ba_data)
bd_summary <- summary(ba_mod)
bd_ktable <- bd_summary$coefficients %>% 
  as_tibble() %>% 
  mutate(Parameter = c("Intercept", "Age (linear)"),
         across(c(Estimate, `Std. Error`, `df`, `t value`), 
                \(x) format_signif(x, digits = 3)),
         `p value` = format_pval(`Pr(>|t|)`, digits = 2),
         Response = "Breeding departure") %>%
  select(-`Pr(>|t|)`) %>% 
  relocate(Response, Parameter)

ma_data <- phenology_models$model_data[[3]]
ma_mod <- lmer(doy ~ age + (1 | animalID) + (1 | year), data = ma_data)
ma_summary <- summary(ma_mod)
ma_ktable <- ma_summary$coefficients %>% 
  as_tibble() %>% 
  mutate(Parameter = c("Intercept", "Age (linear)"),
         across(c(Estimate, `Std. Error`, `df`, `t value`), 
                \(x) format_signif(x, digits = 3)),
         `p value` = format_pval(`Pr(>|t|)`, digits = 2),
         Response = "Molt arrival") %>%
  select(-`Pr(>|t|)`) %>% 
  relocate(Response, Parameter)

md_data <- phenology_models$model_data[[4]]
md_mod <- lmer(doy ~ age + (1 | animalID) + (1 | year), data = md_data)
md_summary <- summary(md_mod)
md_ktable <- md_summary$coefficients %>% 
  as_tibble() %>% 
  mutate(Parameter = c("Intercept", "Age (linear)"),
         across(c(Estimate, `Std. Error`, `df`, `t value`), 
                \(x) format_signif(x, digits = 3)),
         `p value` = format_pval(`Pr(>|t|)`, digits = 2),
         Response = "Molt departure") %>%
  select(-`Pr(>|t|)`) %>% 
  relocate(Response, Parameter)


rbind(ba_ktable, bd_ktable, ma_ktable, md_ktable) %>% 
  knitr::kable() %>% 
  kableExtra::kable_styling() %>% 
  kableExtra::collapse_rows(columns = 1)
```
