---
title: "Elephant Seal Age JAE Intraspecific Variation"
format: 
  html: 
    echo: false 
editor: visual
bibliography: references.bib
---

```{r}
#| label: setup
#| include: false  

library(tidyverse)
library(ggthemes)
library(cowplot)

sealdat <- read_csv(here::here("data/raw/128L pull 2023_06_29.csv"), 
                    show_col_types = FALSE) %>% 
  mutate(observed = if_else(observed == "B", "Breeder", "Non-breeder"))

```

## Introduction

Our hypotheses:

1.  Elephant seal reproductive success declines with age.

2.  Reproductive declines persist to the next generation (maternal effect senescence). 

3.  Seals optimize their annual cycle timing to compensate for reproductive declines.

Note: This paper is similar to another paper about polar bears (*Ursus maritimus*) [@naciri2022].

## Methods

## Results 

### H1: Age distribution and breeding success

We predicted that breeding success would decline with age. We found that breeding success increases to prime age, then decreases at older ages (@fig-age_dist).

```{r}
#| label: fig-age_dist
#| fig-cap: "A: Age structure of northern elephant seals at Año Nuevo. B: Percentage of each age class that were observed breeding."
#| fig-width: 6
#| fig-height: 8

#Age distribution panel A 

age_dist <- sealdat %>% 
  group_by(age) %>% 
  count(age) %>% 
  ggplot(aes(x = age, y = n)) + 
  geom_col(fill = "#829BD1") + 
  labs(x = "Number of individuals", y = "Age (Years)") + 
  theme_few()

#Panel B: percentage of each age class that were breeders or  nonbreeders

breed_nonbreed <- sealdat %>% 
  group_by(age) %>% 
  summarize(n_observed = sum(!is.na(observed)), 
            perc_breed = sum(observed == "Breeder") / n_observed, 
            se = (perc_breed * (1 - perc_breed) / n_observed)^0.5) %>% 
  ggplot(aes(x = age, y = perc_breed)) + 
  stat_smooth(method = "lm", 
              formula = y ~ poly(x , 2), 
              se = TRUE, 
              color = "#849ED5", 
              fill = "#C8D3ED") +
  geom_point() + 
  geom_errorbar(aes(ymin = perc_breed - se, ymax = perc_breed + se), width = 0) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) + 
  labs(x = "Maternal Age (Years)", y = "% Breeders") + 
  theme_few()

plot_grid(age_dist, breed_nonbreed, ncol = 1, labels = "AUTO")
  

```

### H2: Maternal effect senescence 

We predicted that reproductive declines would carry over to the next generation.

We found that as maternal age increases, pup sex skews slightly towards males. Pup survival declines slightly with maternal age, and pup recruitment is not significantly affected by maternal age (@fig-repro_age).

We reject our hypothesis: reproductive declines do not persist to the next generation.

```{r}
#| label: fig-repro_age
#| fig-cap: "Pup sex skews slightly towards males with increasing maternal age. Pup survival decreases slightly, but survival and recruitment are not significantly affected by maternal age."
#| fig-width: 6
#| fig-height: 8

#A: Pup sex shifts towards males with increasing maternal age.

pup_sex <- sealdat %>% 
  filter(pupsex %in% c("M", "F"), 
         age < 18) %>% 
  group_by(age, pupsex) %>%
  count(pupsex) %>% 
  pivot_wider(values_from = n, names_from = pupsex) %>% 
  mutate(num = M + F,
         percf = F / num,
         se = (percf * (1 - percf) / num)^0.5) %>% 
  ggplot(aes(x = age, y = percf, weight = num)) +
  geom_hline(yintercept = 0.5, lty = 2) +
  stat_smooth(method = "glm", 
              method.args = list(family = "binomial"), 
              formula = y ~ x, 
              color = "#829BD1", 
              fill = "#C8D3ED") +
  geom_point() +
  geom_errorbar(aes(ymin = percf - se, ymax = percf + se), width = 0) +
  scale_y_continuous(labels = scales::percent, limits = c(0, NA)) +
  labs(x = "Maternal Age (Years)", y = "% Pups Female") +
  theme_few()

#B Pup survival and recruitment doesn't change with increased maternal age.

pup_surv_recruit <- sealdat %>% 
  mutate(pup_surv = ifelse(pupseeneveragain > 0, 1, 0)) %>% 
  filter(pupsex == "F") %>% 
  group_by(age) %>%
  summarize(n_surv = sum(!is.na(pup_surv)),
            mean_surv = sum(pup_surv, na.rm = TRUE) / n_surv,  
            se_surv = (mean_surv * (1 - mean_surv) / n_surv)^0.5,
            n_recruit = sum(!is.na(puprecruited)), 
            mean_recruit = sum(puprecruited, na.rm = TRUE) / n_recruit, 
            se_recruit = (mean_recruit * (1 - mean_recruit) / n_recruit)^0.5)  %>%  
  pivot_longer(cols = c(mean_surv, se_surv, mean_recruit, se_recruit), 
               names_to = c("stat", "metric"), 
               names_pattern = "(.*)_(.*)") %>% 
  pivot_wider(names_from = "stat", 
              values_from = "value") %>% 
  mutate(low = mean - se, 
         high = mean + se, 
         age = age + ifelse(metric == "surv", -0.1, 0.1), 
         metric = factor(metric, 
                         levels = c("recruit", "surv"), 
                         labels = c("Recruitment", "1st Year Survival"))) %>% 
  filter(n_surv > 5) %>% 
  ggplot(aes(x = age, y = mean, color = metric)) + 
  geom_smooth(aes(weight = n_surv), 
              method = "glm", 
              method.args = list(family = binomial), 
              formula = y ~ x, 
              fill = "#C8D3ED") +
  geom_pointrange(aes(ymin = low, ymax = high)) +
  scale_color_manual(values = c("#829BD1", "#E9B668")) + 
  scale_y_continuous(labels = scales::percent, limits = c(0, .5)) + 
  labs(x = "Maternal Age (Years)", y = "% Pups") +
  theme_few() + 
  theme(legend.position = c(0.1,0.97), 
        legend.justification = c(0,1), 
        legend.title = element_blank(), 
        legend.direction = "horizontal")

plot_grid(pup_sex, pup_surv_recruit, ncol = 1, labels = "AUTO")

```

### H3: Changes to annual cycle timing with increased maternal age

We predicted that seals would optimize the timing of their annual cycle.

We found that as maternal age increases, there are slight adjustments to the timing of life history events. These events (breeding/molting arrival/departure) are typically shifted later for older animals (@fig-annual_timing).

```{r}
#| label: fig-annual_timing
#| fig-cap: Timing of life history events across maternal age classes. 

sealdat %>% 
  mutate(ba = firstobsbreeddoy, 
         bd = firstobsbreeddoy + breeddur, 
         ma = lastobsmoltdoy - moltdur, 
         md = lastobsmoltdoy) %>% 
  drop_na(age, ba, bd, ma, md) %>% 
  group_by(age) %>% 
  summarize(across(ba:md, mean)) %>% 
  pivot_longer(cols = -age, 
               names_to = "event", 
               values_to = "doy") %>% 
  group_by(age) %>% 
  group_modify(\(data, keys) {
    tibble(event = c("Long trip", "Breeding", "Short trip", "Molting", "longtrip2"),
           doy = c(data$doy, 365) - c(0, data$doy))
  }) %>% 
  ungroup() %>% 
  mutate(event = factor(event, levels = c("Long trip", 
                                          "Breeding", 
                                          "Short trip", 
                                          "Molting", 
                                          "longtrip2"))) %>% 
  ggplot(aes(x = age, y = doy, fill = event)) + 
  geom_area(position = position_stack(reverse = TRUE),
            outline.type = "full",
            color = "white") +
  scale_fill_discrete(type = c("#829BD1", "#E9B668", "#7EC6CC", "#E091B9", "#829BD1"),
                      breaks = c("Long trip", "Breeding", "Short trip", "Molting")) + 
  labs(x = "Maternal Age (Years)", y = "Day of Year") + 
  theme_few() + 
  theme(legend.position = "right",
        legend.title = element_blank())

```

## Discussion
